---
title: "Causal Analysis"
author: "Steph Low"
date: "2024-10-17"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tableone)
library(Matching)
library(MatchIt)
library(tidyverse)
library(LaplacesDemon)
```


#### Goal: evaluate the impact of National Supported Work (NSW) Demonstration, which is a labor training program on post-intervention income levels. The interest is in estimating the causal effect of this training program on income.

The Lalonde data set has 614 subjects and 10 variables: 

- `age`: age in years
- `educ`: years of schooling
- `black`: indicator variable for blacks
- `hispan`: indicator variable for Hispanics
- `married`: indicator variable for marital status
- `nodegree`: indicator variable for high school diploma
- `re74`: real earnings in 1974
- `re75`: real earnings in 1975
- `re78`: real earnings in 1978 (the outcome variable; post-intervention income)
- `treat`: indicator variable for treatment status

Potential confounding variables are `age`, `educ`, `black`, `hispan`, `married`, `nodegree`, `re74`, and `re75`. 

### Data Preparation

First I need to prepare the data.

```{r load data}
# load data
data(lalonde)

# view data
# View(lalonde)

# convert indicator variables to numeric and prep data
age <- lalonde$age
educ <- lalonde$educ
re74 <- lalonde$re74
re75 <- lalonde$re75
married <- lalonde$married
nodegree <- lalonde$nodegree
black <- as.numeric(lalonde$race=="black") # 1 for black, 0 otherwise
hispan <- as.numeric(lalonde$race=="hispan") # 1 for hispanic, 0 otherwise
treatment <- lalonde$treat
outcome <- lalonde$re78

# create a dataset with these variables for simplicity
data <- cbind(age, educ, re74, re75, married, nodegree, black, hispan, treatment, outcome)
data <- data.frame(data)

# get covariates
xvars <- c("age", "educ", "black", "hispan", "married", "nodegree", "re74", "re75")
```

### Unmatched Data 

Now that the data is prepared, I want to find the standardized mean differences (SMD) for all the confounding variables prior to matching.


```{r SMD pre-matching}
# create Table1, pre-matching
table1 <- CreateTableOne(vars=xvars             # covariates to be summarized
                         , strata="treatment"   # stratifying the groups by treatment
                         , data=data            
                         , test=FALSE           # don't do groupwise comparisons
                         )

# include the standardized mean difference (SMD)
print(table1, smd=TRUE)
```
Next I want to find the raw mean of real earnings in 1978 (outcome) for treated subjects minus the mean of real earnings in 1978 for untreated subjects.

```{r outcome treatment vs control}
# calculate means by treatment group
raw_means <- aggregate(x=data$outcome
                       , by = list(data$treatment)
                       , FUN = mean)

# calculate the difference between real earnings of the treated subjects minus the untreated subjects
raw_means$x[raw_means$Group.1==1] - raw_means$x[raw_means$Group.1==0]

```

### Propensity Score Estimation

Next I want to fit a propensity score model. I'll estimate the propensity score using logistic regression where the outcome is treatment, including the 8 confounding variables in the model as predictors, with no interaction terms or non-linear terms (such as squared terms).  

The propensity score for each subject is the probability of receiving treatment given the covariates. 


```{r propensity score model}
# fit a propensity score model using logistic regression
psmodel <- glm(treatment~age+educ+re74+re75+married+nodegree+black+hispan
               , family=binomial()
               , data=data
               )

# show model summary
summary(psmodel)

# estimate the propensity scores
pscore <- psmodel$fitted.values

# find the min and max estimated propensity scores
min(pscore)
max(pscore)
```
### Propensity Score Matching

Next I want to match subjects on their propensity scores. 

In this case I will pair match:

- without replacement

- without a maximum distance tolerated for matching (no caliper)

- on the propensity score instead of the logit of the propensity score.

Once the matching is done, I will find the standardized mean differences for the matched data. 

```{R propensity matching}
# set seed for reproducibility
set.seed(931139)

# match on propensity score
psmatch <- Match(Tr=data$treatment     # treatment
                 , M=1                 # pair matching
                 , X=pscore            # variables to match on (estimated propensity scores)
                 , replace=FALSE       # no replacement
                 )

# extracted the matched data
matchedData <- data[unlist(psmatch[c("index.treated", "index.control")]),]

# create Table1, post-matching
matchedTable1 <- CreateTableOne(vars=xvars              # covariates to be summarized
                                 , strata="treatment"   # stratifying the groups by treatment
                                 , data=matchedData            
                                 , test=FALSE           # don't do groupwise comparisons
                                 )

# get SMD
print(matchedTable1, smd=TRUE)
```


Next I'll do the same propensity matching, but this time with a caliper of 0.1 to limit the maximum tolerated distance for matching.

- a common caliper value is `caliper = 0.2 * SD(logit(propensityScore))`

```{r caliper}
# set seed for reproducibility
set.seed(931139)

# match on propensity score
psmatch2 <- Match(Tr=data$treatment    # treatment
                 , M=1                 # pair matching
                 , X=pscore            # variables to match on (estimated propensity scores)
                 , replace=FALSE       # no replacement
                 , caliper=0.1
                 )

# extracted the matched data
matchedData2 <- data[unlist(psmatch2[c("index.treated", "index.control")]),]

# create Table1, post-matching
matchedTable2 <- CreateTableOne(vars=xvars              # covariates to be summarized
                                 , strata="treatment"   # stratifying the groups by treatment
                                 , data=matchedData2            
                                 , test=FALSE           # don't do groupwise comparisons
                                 )

# get SMD
print(matchedTable2, smd=TRUE)
```

I'll again do some propensity matching, but this time with the logit of the propensity score and a caliper.

```{r logit}
# set seed for reproducibility
set.seed(931139)

# match on propensity score
psmatch3 <- Match(Tr=data$treatment    # treatment
                 , M=1                 # pair matching
                 , X=logit(pscore)     # variables to match on (estimated propensity scores)
                 , replace=FALSE       # no replacement
                 , caliper=0.1
                 )

# extracted the matched data
matchedData3 <- data[unlist(psmatch2[c("index.treated", "index.control")]),]

# create Table1, post-matching
matchedTable3 <- CreateTableOne(vars=xvars              # covariates to be summarized
                                 , strata="treatment"   # stratifying the groups by treatment
                                 , data=matchedData3            
                                 , test=FALSE           # don't do groupwise comparisons
                                 )

# get SMD
print(matchedTable3, smd=TRUE)
```
### Outcome Analysis

Next I'll do the outcome analysis.


```{r outcome analysis}
# outcome analysis
y_treatment <- matchedData2$outcome[matchedData2$treatment==1]
y_control <- matchedData2$outcome[matchedData2$treatment==0]

# pairwise difference
diff <- y_treatment - y_control

# mean of real earnings in 1978 for treated subjects minus control subjects
mean(diff)

# paired t test
t.test(diff)

# let's see how this all differs with the logit-matched data

# outcome analysis
y_treatment2 <- matchedData3$outcome[matchedData3$treatment==1]
y_control2 <- matchedData3$outcome[matchedData3$treatment==0]

# pairwise difference
diff2 <- y_treatment2 - y_control2

# mean of real earnings in 1978 for treated subjects minus control subjects
mean(diff2)

# paired t test
t.test(diff2)

```
There wasn't a difference in the outcome based off the different matched sets I did.

With a p-value of `0.1411`, I cannot reject the null hypothesis that there is no treatment effect. The NSW may not have had any impact on post-intervention income levels.

For more reading on this this topic, check out this [example](https://cran.r-project.org/web/packages/tableone/vignettes/smd.html).